---
title: "GSE41762 Regression Analysis"
author: "Cindy Zhang + edit"
date: "2023-02-19"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(GEOquery)
library(limma)
library(ggfortify)
```

### Loading Data

```{r}
eset <- getGEO("GSE41762", getGPL = FALSE)[[1]]
# eset <- getGEO(filename = "~/Downloads/GSE41762_series_matrix.txt.gz", getGPL = FALSE)
head(eset, 10)
str(pData(eset))
head(pData(eset)[1:3,1:5]) %>% kable()
```

### Data Wrangling

```{r}
pData(eset) <- pData(eset) %>%
  mutate(sample_id = geo_accession) %>% 
  select("status:ch1",sample_id,"Sex:ch1", "bmi:ch1", "age:ch1") 
colnames(pData(eset)) <- c("statu","sample_id","sex","bmi","age")

pData(eset)$bmi %>% as.double()
pData(eset)$age %>% as.double()

pData(eset) <- pData(eset) %>% 
  mutate(status = case_when(
         grepl("Non-diabetic donor", statu) ~ "nont2d",
         grepl("Diabetic donor", statu) ~ "t2d")) %>%
  mutate(BMI = case_when(
    bmi >  30 ~ "over30",
    bmi < 30 ~ "below30"
  )) %>% 
  select(-c(bmi,statu))
```

### Arrange factor levels

```{r}
pData(eset) <- pData(eset) %>% 
  mutate(BMI = fct_relevel(BMI, "below30", "over30")) %>% 
  mutate(status = as.factor(status))
```

### match ID between matrixes

```{r}
identical(colnames(exprs(eset)), pData(eset)$sample_id)
```

### Missing Values

-   The missing data are found in sample ID 49 through 77

```{r}
express <- exprs(eset) %>% 
  as.data.frame() 
#samples with missing values 
names(colSums(is.na(express))>0)
express <- na.omit(express)
```

### Combine data

```{r}
# metaData
id <- colnames(express)
MetaData <- pData(eset) %>% 
  select(sample_id, status, BMI) %>% 
  mutate(samples = sample_id) %>% 
  filter(sample_id %in% id)

toLongerMeta <- function(expset) {
    stopifnot(class(expset) == "ExpressionSet")
    expressionMatrix <- longExpressionMatrix <- express %>% 
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    pivot_longer(cols = !gene, 
                 values_to = "Expression",
                 names_to = "sample_id") %>%
    left_join(MetaData)
  return(expressionMatrix)
}

joint <- toLongerMeta(eset)
head(joint, 3) %>% kable()
identical(MetaData$sample_id, colnames(express))

experiment_labels <- c(rep("Experiment 1 Samples", 48), rep("Experiment 2 Samples", 29))
MetaData$experiment = experiment_labels
```

## Perform PCA to analyze batch effect

```{r}
# pca_res_new <- prcomp(express, scale=TRUE)
# autoplot(pca_res_new)
# autoplot(pca_res, loadings.label = TRUE, x = 1, y = 2)
# autoplot(pca_res, loadings = TRUE, loadings.label = TRUE, x = 3, y = 4)

express_scaled <- scale(t(express), center = TRUE, scale = TRUE)
s <- svd(express_scaled)

loadings <- s$v[, 1:3]

scores <- express_scaled %*% loadings

pca_data <- as_tibble(scores) %>% rename(PC1 = V1, PC2 = V2, PC3 = V3) 
pca_data$sample_id = colnames(express)

pca_data <- left_join(pca_data, MetaData, by = "sample_id")

ggplot(pca_data, aes(x=PC2, y=PC1, shape=status, color = BMI)) + geom_point(size=3)
ggplot(pca_data, aes(x=PC3, y=PC2, shape=status, color = BMI)) + geom_point(size=3)
ggplot(pca_data, aes(x=PC3, y=PC1, shape=status, color = BMI)) + geom_point(size=3)

#geom_text(aes(label = sample_id), size=2, nudge_y = 0.05)

# ggplot(pca_data, aes(x=PC1, y=PC2, color = experiment)) + geom_point(size=3)
# ggplot(pca_data, aes(x=PC2, y=PC3, color = experiment)) + geom_point(size=3)
# ggplot(pca_data, aes(x=PC3, y=PC1, color = experiment)) + geom_point(size=3)
```

```{r}
pca_res <- prcomp(t(express), scale=TRUE)


group <- c(rep("A",48), rep("B",29))

meta2 <- MetaData %>%
cbind(.,group)

ggplot(data = pca_res, aes(x = PC1, y=PC2, color = meta2$group)) +
  geom_point(size = 3)

```

-   No discernable batch effect is observed. Thus, can combine thedata from both cohorts for the following study:

# Samples in each group

```{r, echo=FALSE}
table(MetaData$BMI, MetaData$status)
```

### pivot data format

```{r}
toLonger <- function(expressionMatrix) {
    expressionMatrix <- longExpressionMatrix <- expressionMatrix %>% 
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    pivot_longer(cols = !gene, 
                 values_to = "Expression",
                 names_to = "sample_id") 
  return(expressionMatrix)
}

options(repr.plot.width = 30, repr.plot.height =2)

toLonger(express) %>% 
  ggplot(aes(x=sample_id, y= Expression, color=sample_id)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),legend.position = "none") + 
  labs(x = "sample", y = "Gene Expression") 
```

# linear model

```{r}
modm <- model.matrix(~BMI*status, MetaData)
lmFitEb <- eBayes(lmFit(express, modm))
```

### DE genes in obese vs. nonobese healthy samples

```{r}
DEobH <- topTable(lmFitEb, number = Inf, adjust.method="BH", p.value = 0.05, coef= "BMIover30")

DEobH 
```

### DE genes in obese vs. nonobese T2D samples

```{r}
DEObT2d <- topTable(lmFitEb, number = Inf, adjust.method="BH", p.value = 0.05, coef= "BMIover30:statust2d") 
DEObT2d
```

### DE genes in T2D vs. Healthy in nonobese samples

```{r}
degT2dNonOb <- topTable(lmFitEb, number = Inf, adjust.method="BH", p.value = 0.05, coef= "statust2d")

degT2dNonOb
```

### Saving relevant data for aim 2 gene enrichment analysis

```{r}
saveRDS(degT2dNonOb, file = "ObvsNonObHealthy.RDS")
```

-   Note: Result can be loaded into Aim 2 analysis using the following code `readRDS("ObvsNonObHealthy.RDS")`
