---
title: "GSE41762 Regression Analysis"
author: "Cindy Zhang"
date: "2023-03-23"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(GEOquery)
library(limma)
library(ggfortify)
```
### Loading Data
```{r}
eset <- getGEO("GSE41762", getGPL = FALSE)[[1]]
# eset <- getGEO(filename = "~/Downloads/GSE41762_series_matrix.txt.gz", getGPL = FALSE)
head(eset, 10)
str(pData(eset))
head(pData(eset)[1:3,1:5]) %>% kable()
```
### Data Wrangling
```{r}
pData(eset) <- pData(eset) %>%
  mutate(sample_id = geo_accession) %>% 
  select("status:ch1",sample_id,"Sex:ch1", "bmi:ch1", "age:ch1") 
colnames(pData(eset)) <- c("statu","sample_id","sex","bmi","age")
pData(eset)$bmi %>% as.double()
pData(eset)$age %>% as.double()
pData(eset) <- pData(eset) %>% 
  mutate(status = case_when(
         grepl("Non-diabetic donor", statu) ~ "nont2d",
         grepl("Diabetic donor", statu) ~ "t2d")) %>%
  mutate(BMI = case_when(
    bmi >  30 ~ "over30",
    bmi < 30 ~ "below30"
  )) %>% 
  select(-c(bmi,statu))
```
Arrange Factor Levels
```{r}
pData(eset) <- pData(eset) %>% 
  mutate(BMI = fct_relevel(BMI, "below30", "over30")) %>% 
  mutate(status = as.factor(status))
```
Match ID between matrices
```{r}
identical(colnames(exprs(eset)), pData(eset)$sample_id)
```
Missing Values
-   The missing data are found in sample ID 49 through 77
```{r}
express <- exprs(eset) %>% 
  as.data.frame() 
#samples with missing values 
names(colSums(is.na(express))>0)
express <- na.omit(express)
```
Combine Data
```{r}
# metaData
id <- colnames(express)
MetaData <- pData(eset) %>% 
  select(sample_id, status, BMI) %>% 
  mutate(samples = sample_id) %>% 
  filter(sample_id %in% id)
toLongerMeta <- function(expset) {
    stopifnot(class(expset) == "ExpressionSet")
    expressionMatrix <- longExpressionMatrix <- express %>% 
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    pivot_longer(cols = !gene, 
                 values_to = "Expression",
                 names_to = "sample_id") %>%
    left_join(MetaData)
  return(expressionMatrix)
}
joint <- toLongerMeta(eset)
head(joint, 3) %>% kable()
identical(MetaData$sample_id, colnames(express))
experiment_labels <- c(rep("Experiment 1 Samples", 48), rep("Experiment 2 Samples", 29))
MetaData$experiment = experiment_labels
```
The first 48 samples were from one experiment cohort while the remaining 29 samples were from another. The batch effect between the two cohorts were evaluated   
```{r}
express_scaled <- scale(t(express), center = TRUE, scale = TRUE)
s <- svd(express_scaled)
loadings <- s$v[, 1:3]
scores <- express_scaled %*% loadings
svd <- as_data_frame(scores) %>% rename(u1 = V1, u2 = V2, u3 = V3) 
svd$sample_id = colnames(express)
svd <- left_join(svd, MetaData, by = "sample_id")

ggplot(svd, aes(x=u1, y=u2, color = experiment)) + geom_point(size=3)
ggplot(svd, aes(x=u2, y=u3, color = experiment)) + geom_point(size=3)
ggplot(svd, aes(x=u3, y=u1, color = experiment)) + geom_point(size=3)
```

- A moderate batch effect was observed for u2 vs. u1, the first 48 samples will be used for downstream analysis 
```{r}
meta <- pData(eset)[1:48,]
```
Samples in Each Group
- There are insufficient sample to fit an model with BMI as a covariate. Thus, a model will be focused on ~ T2D status
```{r, echo=FALSE}
table(meta$BMI, meta$status)
```
Pivot Data Format
```{r}
toLonger <- function(expressionMatrix) {
    expressionMatrix <- longExpressionMatrix <- expressionMatrix %>% 
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    pivot_longer(cols = !gene, 
                 values_to = "Expression",
                 names_to = "sample_id") 
  return(expressionMatrix)
}
options(repr.plot.width = 30, repr.plot.height =2)
toLonger(express[1:48]) %>% 
  ggplot(aes(x=sample_id, y= Expression, color=sample_id)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),legend.position = "none") + 
  labs(x = "sample", y = "Gene Expression") 
```

### Linear Model Fitting
```{r}
modm <- model.matrix(~status, meta)
lmFitEb <- eBayes(lmFit(express[1:48], modm))
```
Significant DE genes in T2D vs. Healthy
```{r}
degT2d  <- topTable(lmFitEb, number = Inf, adjust.method="BH", p.value = 0.05, coef= "statust2d")
degT2d %>% head(10) 
```
Saving all DE gene for aim 2 gene set enrichment analysis
```{r}
degT2d_DE  <- topTable(lmFitEb, number = Inf, coef= "statust2d")
saveRDS(degT2d_DE, file = "degT2d_DE.RDS")
```
Note: Result can be loaded into Aim 2 analysis using the following code `readRDS("degT2d_DE")`
